---
title: "Homework 1"
author: "Natalie Sawtelle"
date: "2025-09-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggfortify)
library(knitr)
library(emmeans)
```

# Problem 1: Chick Weights 

``` {r  task 1: anova model in grams and ounces}
chick_weights <- chickwts |>
  mutate(weight_oz = weight/28.35)
chick_anova_g <- aov(weight ~ feed, data=chick_weights)
chick_anova_oz <- aov(weight_oz ~ feed, data=chick_weights)
```

``` {r task 2: anova tables}
summary(chick_anova_g)
summary(chick_anova_oz)
```

**Both tests result in the same F and P-values, this shows that unit conversion does not affect the results of hypothesis testing.**

``` {r task 3: equal variance}
autoplot(chick_anova_g)
```

**The spread of residuals looks uneven across the groups, this suggests possible violations of the equal variance assumption.**

```{r task 4: constant variance assumption}
chick_log_anova <- aov(log10(weight) ~ feed, data=chick_weights)
autoplot(chick_log_anova)
```

**The Residuals vs Fitted and Scale-Location plots show uniform spread. This indicates that the log-transformation improves the constant variance assumption compared to the not transformed model.**

``` {r task 5: ANOVA table for the model fit with the logarithm of the chick weights as a response?????}
summary(chick_log_anova)
```

**Both conclude that Feed type significantly affects chick weight. Task two gives out large sums of squares, while Task 5 gives out small sums of squares.**

**task 6: Transforming the response variable makes the analysis more statistically valid by meeting model assumptions, but it does this at the cost of the ability to easily interpret the results.**


#Problem 2:Popcorn

* What are the **Experimental Units** in this study? 
     - **Bags of kernels**
* What is the **response variable** in this study? 
     - **number of kernels poped**
* What is the **factor** in this study? 
     - **oil amount**
* How many **factor levels** are there? 
     - **none, 1/2 teaspoon ,1 teaspoon**
* What are the **treatments**?  
     - **Popcorn popped with no oil, with 1/2 teaspoon of oil, and with 1 teaspoon of oil**
     
```{r}
Popcorn_study <- read.csv("popcorn.csv")
```
     
```{r data cleaning popcorn}
Popcorn_study_ordered <- Popcorn_study |> 
  arrange(desc(Oil), Kernels)
```


```{r ANOVA and residual analysis}
anova_model <- aov(Kernels ~ Oil, data = Popcorn_study)
autoplot(anova_model)
```

**Since the p-value is much smaller than 0.05, we reject the null hypothesis that oil has no effect.**

```{r numberical summary}
summary(anova_model)
```

**The Residuals vs Fitted plot suggests no violation and the Normal Q-Q plot is approximately normal.**


# Problem 3: Ant Pesticide

* What are the **Experimental Units** in this study? 
     - **Colonies**
* What is the **response variable** in this study? 
     - **Number of ants killed**
* What is the **factor** in this study? 
     - **Pesticide**
* How many **factor levels** are there? 
     - **types A, B, C, and D**
* What are the **treatments**?  
     - **Colonies treated with pesticide type A, type B, type C, or type D**
     
```{r}
ant_pesticide_study <- read.csv("ant_pesticide.csv")
```
     
```{r data cleaning ants}
ant_pesticide_study_ordered <- ant_pesticide_study |>
  mutate(Pesticide = factor(Pesticide)) |>
  arrange(Pesticide, DeadAnts)
```
     
```{r EDA}
ggplot(ant_pesticide_study_ordered) + 
  geom_boxplot(aes(x=DeadAnts, y=Pesticide)) + 
  theme_bw()
```

```{r ANOVA ants}
ant_anova <- aov(DeadAnts ~ Pesticide, data = ant_pesticide_study_ordered)
summary(ant_anova)
```

```{r resdiduals}
autoplot(ant_anova)
```

**Based on this study, there is no significant differences between pesticides type A, B, C, and D. However, there are some indications that types A and D are better than types B and C.**











